---
title: "Lab 2: Use of Deadly Force by Police"
author: "Matt Robinson & Vincent Lee"
date: "9/26/2016"
output: html_document
---

## R Markdown

The following lines of code import necessary libraries to execute this code:

### Preparation / Imports
``` {r}
library(devtools)
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

``` {r eval = F}
install.packages("devtools")
install.packages("flexdashboard")
install.packages("lubridate")
install.packages("dplyr")
install.packages("ggplot2")
devtools::install_github("lionel-/ggstance")
install.packages("ggalt")
```

```{r message = F}
library(flexdashboard)
library(lubridate)
library(dplyr)
library(ggplot2)
library(ggstance)
library(ggalt)
```

## Introduction

For this lab, we took a look at the use of deadly force by police officers in the United States. This analysis is based on the work of Julia Silge of "data flex-dashboard". 

- Link to Article: <https://www.r-bloggers.com/fatal-police-shootings-across-the-u-s/>
- Link to Data: <https://github.com/washingtonpost/data-police-shootings>
- Link to the Dashboard Visualization: <https://gist.github.com/juliasilge/9acbe97c549502bac85404779edceba0>
- Link to the Washington Post Visualization <https://www.washingtonpost.com/graphics/national/police-shootings/>

The data we imported is a compilation of every fatal shooting in the United States by a police officer in the line of duty from a database organized by the Washington Post since Jan. 1, 2015.

Since 2015, The Washington Post has recorded dozens of details about each killing:

- **id**: unique id of the person
- **name**: first, middle, and last name of the person
- **date** the date of the fatal shooting  
- **manner_of_death**: description of whether they were Beaten or Shot
- **armed**: whether the person killed was armed 
- **age**: person's age
- **gender**: Male or Female
- **race**: includes the race of the deceased
- **city**: location of city area
- **state**: location of state
- **threat level**: the circumstances of the shooting
- **signs_of_mental_illness**: whether the victim was experiencing a mental-health crisis 
- **flee**: were they running away by Foot, Not Fleeing, Car, Other
- **body_camera**: was the victim recorded by a body camera

In 2016, The Washington post went to gather additional information about each fatal shooting that occurred during this year. There were reports, websites, and social media that have been monitoring independent databases about fatal police shootings. More than a dozen additional details were collected about the officers in each shooting. 

We want to look at the data and make an in-depth analysis on the different demographic, date, location, and circumstance of death within Fatal Police Shootings Data.

## Analysis

We imported the data set obtained from the Washington Post police shootings database. Then, we clean up the code by changing dates, genders, races, and manners of death to an understandable format.
```{r}
fatalshootings <- read.csv("./fatal-police-shootings-data.csv", stringsAsFactors = FALSE)
fatalshootings$date <- ymd(fatalshootings$date)
fatalshootings[,c(4:5,7:8,9:14)] <- lapply(fatalshootings[,c(4:5,7:8,9:14)], as.factor)
levels(fatalshootings$gender) <- c("Female", "Male")
levels(fatalshootings$race) <- c("Unknown", "Asian", "Black", "Hispanic", "Unknown", "Other", "White")
fatalshootings$race <- factor(fatalshootings$race, levels(fatalshootings$race)[c(5,2,1,4,3,6)])
levels(fatalshootings$flee) <- c("Unknown", "Car", "Foot", "Not fleeing", "Other")
fatalshootings$flee <- factor(fatalshootings$flee, levels(fatalshootings$flee)[c(1,5,3,2,4)])
levels(fatalshootings$manner_of_death) <- c("Beaten", "Shot", "Shot/Tasered")
fatalshootings$manner_of_death <- factor(fatalshootings$manner_of_death, levels(fatalshootings$manner_of_death)[c(1,3,2)])
```

Once we have our clean data, we are ready to start creating interesting visualizations using ggplot.

### Demographic Analysis

#### Deaths by Race/Ethnicity
```{r}
ggplot(data = fatalshootings, aes(y = race)) + 
        geom_barh(aes(fill = ..count..)) +
        theme_minimal(base_size = 13) +
        theme(legend.position = "none") +
        scale_x_continuous(expand=c(0,0)) +
        scale_fill_gradient(low = "royalblue3", high = "navyblue") +
        labs(y = NULL, x = "Number of deaths")
```

> This is a simple graph yet you can scrutinize a lot information from it. Instead of the typical vertical graph the original creator Julia decided to make the graph horizontal. Based on the number of Fatal Shootings she also implemented a lighter-to-darker gradient transition. Overall, she makes good use of whitespace and direction. The same goes with ink to data ratio. The most important feature is that one can clearly tell which ethinicity suffers the most from fatal police shootings. Our only question is what categorizes the unknown.

#### Deaths by Gender
```{r}
ggplot(data = fatalshootings, aes(y = gender)) + 
        geom_barh(aes(fill = ..count..)) +
        theme_minimal(base_size = 13) +
        theme(legend.position = "none") +
        scale_x_continuous(expand=c(0,0)) +
        scale_fill_gradient(low = "royalblue3", high = "navyblue") +
        labs(y = NULL, x = "Number of deaths")
```

> Clearly, male victims fall into the greater number of fatal police shootings.

#### Deaths by Age
```{r warning = F}
ggplot(data = fatalshootings, aes(x = age)) + 
        geom_histogram(aes(fill = ..count..), bins = 20) +
        theme_minimal(base_size = 13) +
        theme(legend.position = "none") +
        scale_x_continuous(expand=c(0,0)) +
        scale_fill_gradient(low = "royalblue3", high = "navyblue") +
        labs(x = "Age at death", y = "Number of deaths")
```

> It forms a skewed-right distribution. The deaths seem to be the greatest at the age of 30. We are also glad that children are not usually involved wth fatal shootings. Creator uses consistent themes and a clear number of bins for her ggplot.

####Deaths based on mental illness
```{r}
ggplot(data = fatalshootings, aes(y = signs_of_mental_illness)) + 
        geom_barh(aes(fill = ..count..)) +
        theme_minimal(base_size = 13) +
        theme(legend.position = "none") +
        scale_x_continuous(expand=c(0,0)) +
        scale_fill_gradient(low = "royalblue3", high = "navyblue") +
        labs(y = NULL, x = "Number of deaths")
```

```{r}
# I would add the percentages or proportions onto the graph
summary(fatalshootings$signs_of_mental_illness)
(420/1698)
```

> A significant proportion of death is based on mental illness. Could there be some bias in the data collection? What considers to be a mental illness? 

### Date of Shootings

#### Overall

```{r}
ggplot(data = fatalshootings, aes(x = date)) + 
        geom_histogram(aes(fill = ..count..), bins = 25) +
        theme_minimal() +
        theme(legend.position = "none") +
        scale_fill_gradient(low = "royalblue3", high = "navyblue") +
        labs(x = NULL, y = "Number of deaths")
```

#### Deaths by Month
```{r}
ggplot(data = fatalshootings, aes(x = month(date, label = TRUE))) + 
        geom_bar(aes(fill = ..count..)) +
        theme_minimal() +
        theme(legend.position = "none") +
        scale_y_continuous(expand=c(0,0)) +
        scale_fill_gradient(low = "royalblue3", high = "navyblue") +
        labs(x = NULL, y = "Number of deaths")
```

#### Deaths by Day of Week
```{r}
ggplot(data = fatalshootings, aes(x = wday(date, label = TRUE))) + 
        geom_bar(aes(fill = ..count..)) +
        theme_minimal() +
        theme(legend.position = "none") +
        scale_y_continuous(expand=c(0,0)) +
        scale_fill_gradient(low = "royalblue3", high = "navyblue") +
        labs(x = NULL, y = "Number of deaths")
```

> They're the same graph, but different labels and count for the dates. When creating the flexboard, it's good that there is a consistent theme. The creator continues the trend of lighter-to-darker gradient. I also think it was a good call to minimize it to the appropriate number of bins.


### Top 15 States For Fatal Police Shootings

### Location Breakdown

#### Deaths by State
```{r message = F}
stateinfo <- fatalshootings %>% group_by(state) %>% summarise(n = n()) %>% 
        arrange(desc(n)) %>% top_n(15) %>% 
        mutate(state = factor(state, levels = rev(unique(state))))
ggplot(stateinfo, aes(x = n, y = state)) +
        geom_barh(stat="identity", aes(fill = n)) +
        #geom_stateface(aes(y=state, x=7, label=as.character(state)), colour="white", size=8) +
        geom_text(aes(x = 17, y = state, label=as.character(state)), color="white", size=4) +
        labs(y = NULL, x = "Number of deaths") +
        scale_fill_gradient(low = "royalblue3", high = "navyblue") +
        theme_minimal(base_size = 13) +
        theme(axis.text.y=element_blank()) +
        theme(legend.position = "none") +
        scale_x_continuous(expand=c(0,0))
```

> This graph is a little bit more complicated than the rest. The creator focuses on seperating the top 15 states when finding the distinct variables. Then the creator maps it to the barplot. Original there were state icons, but we couldn't find the 'geom_stateface' asset. 

### Was The Person Killed Armed

#### Was the victim armed?
```{r message = F}
armedinfo <- fatalshootings %>% group_by(armed) %>% summarise(n = n()) %>% 
        arrange(desc(n)) %>% top_n(10) %>% 
        mutate(armed = factor(armed, levels = rev(unique(armed))))

ggplot(data = armedinfo, aes(x = n, y = armed)) + 
        geom_barh(stat="identity", aes(fill = n)) +
        theme_minimal(base_size = 13) +
        theme(legend.position = "none") +
        scale_x_continuous(expand=c(0,0)) +
        scale_fill_gradient(low = "royalblue3", high = "navyblue") +
        labs(y = NULL, x = "Number of deaths")
```

> There was a great number of unusually recored weapons. It makes sense that most fatal shootings would occur with the victim holding out a gun. 

### Circumstances of Death

#### Was the officer wearing a body camera?
```{r}
ggplot(data = fatalshootings, aes(y = body_camera)) + 
        geom_barh(aes(fill = ..count..)) +
        theme_minimal(base_size = 13) +
        theme(legend.position = "none") +
        scale_x_continuous(expand=c(0,0)) +
        scale_fill_gradient(low = "royalblue3", high = "navyblue") +
        labs(y = NULL, x = "Number of deaths")
```

> There should be more officers wearing a body camera. 

#### Was the victim fleeing?
```{r}
ggplot(data = fatalshootings, aes(y = flee)) + 
        geom_barh(aes(fill = ..count..)) +
        theme_minimal(base_size = 13) +
        theme(legend.position = "none") +
        scale_x_continuous(expand=c(0,0)) +
        scale_fill_gradient(low = "royalblue3", high = "navyblue") +
        labs(y = NULL, x = "Number of deaths")
```

> Great visualization that shows the amount of victims fleeing or not fleeing. Again, how do you 'categorize' unknown and still have 'other'.

#### How was the person killed?
```{r}
ggplot(data = fatalshootings, aes(y = manner_of_death)) + 
        geom_barh(aes(fill = ..count..)) +
        theme_minimal(base_size = 13) +
        theme(legend.position = "none") +
        scale_x_continuous(expand=c(0,0)) +
        scale_fill_gradient(low = "royalblue3", high = "navyblue") +
        labs(y = NULL, x = "Number of deaths")
```


> We did not expect that most victims were beaten to death rather than shot to death.

### Follow Up

To enhance the article, we added some more detail to the location graph to have the data represent death per 1 million residents. The population data was retrieved from the United States Census Bureau here <https://www.census.gov/popest/data/datasets.html>. We do some importing, then merge data to perform the calculation, and then plot the new variable, dPerCapita. As you can see, California proportionally does not have the most deaths, but rather New Mexico. Also, Wyoming and Arkansas, who were not in the original visualization are second and third in deaths per million.

#### Death per Capita Breakdown
```{r message=FALSE}
stateinfo <- fatalshootings %>% group_by(state) %>% summarise(n = n()) %>% 
  arrange(desc(n)) %>% 
  mutate(state = factor(state, levels = rev(unique(state))))

census <- read.csv('./NST-EST2015-alldata.csv', stringsAsFactors = FALSE)
census <- transform(census, state = state.abb[match(census$NAME,state.name)])
stateinfo <- merge(stateinfo,census,by="state")
stateinfo$dPerCap <- stateinfo$n / (stateinfo$POPESTIMATE2015 / 1000000)
stateinfo <- stateinfo %>% arrange(desc(dPerCap)) %>% top_n(15) 

ggplot(stateinfo, aes(x = dPerCap, y = state)) +
  geom_barh(stat="identity", aes(fill = dPerCap)) +
  geom_text(aes(x = 20, y = state, label=as.character(state)), color="white", size=4) +
  labs(y = NULL, x = "Number of deaths per 1M people") +
  scale_fill_gradient(low = "royalblue3", high = "navyblue") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") +
  scale_x_continuous(expand=c(0,0))

```

## Follow-Up (Vincent)

Our twist on the fatal police shootings data.

Import the csv file.
```{r}
fatalshootings = read.csv("fatal-police-shootings-data.csv")
```

Simple analysis without making graphical models (throw it into a simple barplot if you do want to make a graphical model)
```{r}
table(fatalshootings$armed)
table(fatalshootings$manner_of_death)
table(fatalshootings$age)
table(fatalshootings$gender)
table(fatalshootings$race)
table(fatalshootings$state)
table(fatalshootings$signs_of_mental_illness)
table(fatalshootings$threat_level)
table(fatalshootings$flee)
table(fatalshootings$body_camera)
```

### Data Visualization.

Creating a count for fatalshootings per day

```{r}
deathsperday = fatalshootings %>%
  group_by(date) %>%
  mutate(count = n())
```


#### Random Linear Model of Fatal Police Shootings Per Day (this is such a bad graph, we are not using this)

```{r}
ggplot(data = deathsperday, aes(x = date, y = count)) + 
  geom_line(aes(colour = count)) +
   stat_smooth(method = "lm", col = "red") +
  scale_colour_gradient2(low = "blue", mid = "green" , high = "red", midpoint = 4) +
        labs(x = NULL, y = "Number of deaths")
```


## Conclusion

- We could've definitely looked into using other variables for supporting our Data Analysis. 
- There could have been issues or bias with the data provided. However all those are caveats for the importance of data collection and analysis.
- If we had time, the stretch goal was to create a graphical map visualization or an interactive visualization where people can hover over and see the data.
